Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-09-08 paper/paper.tex
Minor

\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
\addresses{\textbf{Fan Cheng}\newline
Monash University
\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
\textbf{Anastasios Panagiotelis}\newline
University of Sydney
\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
\textbf{Rob J Hyndman}\newline
Monash University
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
}

@@

@@

For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}. The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}. In this paper, we focus on the multivariate kernel density estimation.

@@

>>>>

\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
\addresses{\textbf{Fan Cheng}\newline
Monash University
\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
\textbf{Anastasios Panagiotelis}\newline
The University of Sydney
\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
\textbf{Rob J Hyndman}\newline
Monash University
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
}

@@

% Adjust headwidth in case user has changed geometry in header-includes
\renewcommand{\headwidth}{\textwidth}

@@

For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][]{Heidenreich2013-bl}[ for details]{Scott2015-vl}.
For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][]{Jones1992-ef,Cao1994-st,Jones1996-cb}[ for reviews]{Wand1994-xu}. The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}. In this paper, we focus on the multivariate kernel density estimation.

@@

\hypertarget{s1simulation}{%
\section{\texorpdfstring{Appendix: Simulation of \(S^1\) for manifold density estimation}{Appendix: Simulation of S\^{}1 for manifold density estimation}}\label{s1simulation}}

To start with, we demonstrate the use of the proposed DC-KDE on a \(1\)-dimensional example. For a \(S^1\) curve centered at \(O\) with radius \(1\), the coordinated of points on the curve could be expressed in polar coordinates \((1, \alpha)\) or Cartesian coordinates \((x, \sqrt{1-x^2})\).

The first simulation starts with a \(1\)-dimensional example.

Manifold: \(S^1\) with center 0 and radius \(r=1\)

Polar coordinates: \((r, \theta)\)

Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).

Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)

Density of \(\theta\): \(f(\theta) = 2 / \pi\)

Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)

The CDF of X is given by
\[
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
\]
Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
\]

Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
and compare it with its true density \(f(x)\).
We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
\]
where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).

Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\). Therefore, the geodesic distance is
\[
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
\]
As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).

{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}

The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
\[
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
\]
is smooth in a local neighborhood \(U\) of \(M\).

Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
\[
\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
\]
Therefore, the estimator for \(p \in S^1\) is given by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
\]
where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.

\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}

\hypertarget{example-of-s2-in-henry2009-om.}{%
\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}

Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
\[
\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
\]
if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
\]
\[
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
\]
The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
Therefore, we obtain that the volume density on the sphere is
\[
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
\]
for \(q \neq p, -p\), and \(\theta_p(p) = 1\).

\hypertarget{s1}{%
\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}

Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
\[
\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
\]
and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
\]
So the riemannian metric is \(g_q = \cos(r)\) and further,
\[
\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
\]
where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).
